Multiple Comparisons in Toolmark Evidence

### 2023-11-21 LPR/index.qmd
compare wire to blade cut, not the other way round

One of the common suggestions in the 2016 PCAST report across pattern disciplines in forensic science was the importance of creating huge databases of e.g. known prints and firearms to support research into matches and closest non-matches[@pcast, fingerprints pg 11, 88, firearms pg 12]. As researchers, we concur with this suggestion, as it is critical for developing objective algorithms and creating useful black-box studies. We have spent considerable time assembling large data sets of tool marks, bullets, and cartridge cases to support our research as well as create public resources to accelerate future research [@csafeDataPortal]. However, we are wary about the implications of database searches across forensic disciplines, for reasons also mentioned in the PCAST report: the necessity of understanding the random match probability for searches against these databases, in particular when using degraded evidence often found in crime scenes[@pcast, pg 52].

This issue has been raised in the past with regard to DNA [@thompson2003] and fingerprints [@fine2006]; one of the many issues identified in the aftermath of the 2004 Madrid bombing case was that the IAFIS database used to locate similar prints is extremely large and thus it is possible for the database to locate unusually similar non-matches. This issue will almost certainly become an issue in other pattern disciplines as databases of firearms evidence such as NIBIN @nibin2021 become more commonly utilized by law enforcement. In this paper, however, we examine a more subtle issue present in tool mark examinations: searches within evidence collected as part of a single case. We begin with a hypothetical set of three scenarios involving the collection of tool mark evidence and then assess the statistical issues involved with each scenario and explore possible resolutions to those issues.

@@

When an examiner conducts a toolmark assessment of wire cut evidence, they are typically working with a tool recovered from the suspect as well as a wire recovered from the scene. The examiner will typically take the tool and use it to cut a piece of metal, known as a blade cut, in order to record the markings of the entire cutting surface simultaneously. This process may be repeated at multiple angles in order to record the effect of different contact surfaces on the resulting striations. The resulting blade cut will be compared to the wire using either visual inspection or perhaps in the future a numerical comparison algorithm similar to those developed for bullet striations [@hare2017automatic] and cartridge cases [@zemmels2023].

>>>>

One of the common suggestions in the 2016 PCAST report across pattern disciplines in forensic science was the importance of creating huge databases of e.g. known prints and firearms to support research into matches and closest non-matches[@pcast, fingerprints pg 11, 88, firearms pg 12]. As researchers, we concur with this suggestion, as it is critical for developing objective algorithms and creating useful black-box studies. We have spent considerable time assembling large data sets of tool marks, bullets, and cartridge cases to support our research as well as create public resources to accelerate future research [@csafeDataPortal]. However, we are wary about the implications of database searches across forensic disciplines, for reasons also mentioned in the PCAST report: the necessity of understanding the random match probability for searches against these databases, in particular when using degraded evidence often found in crime scenes [@pcast, pg 52].

This issue has been raised in the past with regard to DNA [@thompson2003] and fingerprints [@fine2006]; one of the many issues identified in the aftermath of the 2004 Madrid bombing case was that the IAFIS database used to locate similar prints is extremely large and thus it is possible for the database to locate 'unusually' similar non-matches. This issue will almost certainly become an issue in other pattern disciplines as databases of firearms evidence such as NIBIN @nibin2021 become more commonly utilized by law enforcement. In this paper, however, we examine a more subtle issue present in tool mark examinations: searches within evidence collected as part of a single case. We begin with a hypothetical set of three scenarios involving the collection of tool mark evidence and then assess the statistical issues involved with each scenario and explore possible resolutions to those issues.

@@

When an examiner conducts a toolmark assessment of wire cut evidence, they are typically working with a tool recovered from the suspect as well as a wire recovered from the scene. The examiner will typically take the tool and use it to cut a piece of metal, known as a blade cut, to record the markings of the entire cutting surface simultaneously. This process may be repeated at multiple angles in order to record the effect of different contact surfaces on the resulting striations. Markings on the wire will then be compared to these blade cuts using either visual inspection or perhaps in the future a numerical comparison algorithm similar to those developed for bullet striations [@hare2017automatic] and cartridge cases [@zemmels2023]. \hh{let's not just cite CSAFE we are not the only ones. }
